Toronto's Covid-19 Impact: Effect of Age and Gender on Outcomes

### 2024-09-24 paper/paper.qmd
updates to data, paper,

### Table 1: Sample of Raw Covid-19 Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

@@

@@

### Table 2: Sample of Cleaned Covid-19 Data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Cleaned Data Table (Table 2)
cleaned_covid_data %>%
  head(5) %>%
  kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
        booktabs = TRUE)
```

## Summary Statistics

```{r, echo=FALSE, warning=FALSE, message=FALSE}
cleaned_covid_data %>%
  summarise(
    total_cases = n(),
    age_group_distribution = table(age_group),
    gender_distribution = table(client_gender),
    outcome_distribution = table(outcome),
    earliest_reported_date = min(reported_date, na.rm = TRUE),
    latest_reported_date = max(reported_date, na.rm = TRUE)
  )
```
## A Note on Measurement

@@

@@

## Gender vs Outcomes

# Discussion {#sec-discussion}

>>>>

```{r}
#| label: tbl-raw1
#| tbl-cap: "COVID-19 Case Raw Data"
#| echo: false
#| warning: false
#| message: false

library(kableExtra)

@@

```{r}
#| label: tbl-raw2
#| tbl-cap: "COVID-19 Case Raw Data Part 2"
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-clean
#| tbl-cap: "COVID-19 Case Cleaned Data Part 2"
#| echo: false
#| warning: false
#| message: false
#| 
# Cleaned Data Table (Table 2)
cleaned_covid_data %>%
  head(5) %>%
  kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
        booktabs = TRUE)
```

## A Note on Measurement

@@

@fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024.

```{r}
#| label: fig-cumulativecases
#| fig-cap: "COVID-19 Case Progression Over Time, Toronto"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Step 1: Ensure reported_date is in Date format
cleaned_covid_data$reported_date <- as.Date(cleaned_covid_data$reported_date)

# Step 3: Plot cumulative cases over time
ggplot(cleaned_covid_data, aes(x = reported_date, y = x_id)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Cumulative COVID-19 Cases Over Time in Toronto",
       x = "Date",
       y = "COVID-19 Cases") +
  scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))

```{r}
#| label: fig-treatmentoutcomes
#| fig-cap: "Level of Hospitalisation vs Mortality Rates"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Step 1: Calculate mortality rate for intubated, hospitalized, and ICU patients
# Create a summary table with mortality rates

calc_mortality_rate <- function(df, condition_col) {
  total_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes"))
  fatal_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes" & outcome == "FATAL"))
  if (total_patients > 0) {
    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
  } else {
    return(NA)  # Return NA if no patients in the group
  }
}

# Apply the function for each group
hospitalised_mortality_data <- data.frame(
  Condition = c("Intubated", "Hospitalized", "ICU"),
  Mortality_Rate = c(
    calc_mortality_rate(cleaned_covid_data, "ever_intubated"),
    calc_mortality_rate(cleaned_covid_data, "ever_hospitalized"),
    calc_mortality_rate(cleaned_covid_data, "ever_in_icu")
  )
)

# Step 2: Plot the mortality rate as a bar graph
ggplot(hospitalised_mortality_data, aes(x = Condition, y = Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mortality Rate by Condition (Intubated, Hospitalized, ICU)",
       x = "Condition",
       y = "Mortality Rate (%)") +
  theme_minimal() +
  ylim(0, 100)  # Set y-axis limits to 0-100 for percentage

@@

```{r}
#| label: fig-agegroupmortality
#| fig-cap: "Age Group vs Mortality Rates"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Step 1: Define the calc_mortality_rate function for age group
calc_mortality_rate_age <- function(df, age_group_col) {
  total_patients <- nrow(df %>% filter(!!sym(age_group_col) != ""))
  fatal_patients <- nrow(df %>% filter(!!sym(age_group_col) != "" & outcome == "FATAL"))
  if (total_patients > 0) {
    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
  } else {
    return(NA)  # Return NA if no patients in the group
  }
}

# Step 2: Calculate mortality rate for each age group
age_groups <- c("19 and younger", "20 to 29 Years", "30 to 39 Years", 
                "40 to 49 Years", "50 to 59 Years", "60 to 69 Years", 
                "70 to 79 Years", "80 to 89 Years", "90 and older")

mortality_by_age_group <- data.frame(
  Age_Group = age_groups,
  Mortality_Rate = sapply(age_groups, function(group) {
    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
    calc_mortality_rate_age(df_age_group, "age_group")
  })
)

# Step 3: Plot Age Group vs Mortality Rate as a bar chart
ggplot(mortality_by_age_group, aes(x = Age_Group, y = Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mortality Rate by Age Group",
       x = "Age Group",
       y = "Mortality Rate (%)") +
  theme_minimal() +
  ylim(0, 15) +  # Set y-axis limits to 0-15 for percentage
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```{r}
#| label: fig-agegrouphospitalisation
#| fig-cap: "Chance of Hospitalisation by Age Group"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Step 1: Define a function to calculate rates for each condition (hospitalization, intubation, ICU)
calc_condition_rate <- function(df, condition_col) {
  total_patients <- nrow(df %>% filter(!is.na(!!sym(condition_col))))  # Non-missing values for the condition
  condition_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes"))
  if (total_patients > 0) {
    return(condition_patients / total_patients * 100)  # Return rate as a percentage
  } else {
    return(NA)  # Return NA if no patients in the group
  }
}

# Step 2: Calculate the rates for hospitalization, intubation, and ICU admission by age group
age_groups <- c("19 and younger", "20 to 29 Years", "30 to 39 Years", 
                "40 to 49 Years", "50 to 59 Years", "60 to 69 Years", 
                "70 to 79 Years", "80 to 89 Years", "90 and older")

hospitalisation_data <- data.frame(
  Age_Group = age_groups,
  Hospitalization_Rate = sapply(age_groups, function(group) {
    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
    calc_condition_rate(df_age_group, "ever_hospitalized")
  }),
  Intubation_Rate = sapply(age_groups, function(group) {
    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
    calc_condition_rate(df_age_group, "ever_intubated")
  }),
  ICU_Rate = sapply(age_groups, function(group) {
    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
    calc_condition_rate(df_age_group, "ever_in_icu")
  })
)

# Step 3: Reshape the data to long format for grouped bar plot
long_hosp_data <- hospitalisation_data %>%
  pivot_longer(cols = c(Hospitalization_Rate, Intubation_Rate, ICU_Rate),
               names_to = "Condition",
               values_to = "Rate")

# Step 4: Plot Age Group vs Chance of Hospitaliation, Intubation, or ICU as a grouped bar chart
ggplot(long_hosp_data, aes(x = Age_Group, y = Rate, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalisation, Intubation, ICU Admission by Age Group",
       x = "Age Group",
       y = "Percentage (%)") +
  theme_minimal() +
  ylim(0, 25) +  # Set y-axis limits to 0-100%
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```
## Gender vs Outcomes

```{r}
#| label: fig-gendermortality
#| fig-cap: "COVID-19 Mortality by Gender"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Step 1: Define the calc_mortality_rate function (as used previously)
calc_mortality_rate_gender <- function(df, gender_col) {
  total_patients <- nrow(df %>% filter(!!sym(gender_col) != ""))
  fatal_patients <- nrow(df %>% filter(!!sym(gender_col) != "" & outcome == "FATAL"))
  if (total_patients > 0) {
    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
  } else {
    return(NA)  # Return NA if no patients in the group
  }
}

# Step 2: Calculate mortality rate for each gender (MALE, FEMALE, OTHER)
genders <- c("MALE", "FEMALE", "OTHER")

mortality_by_gender <- data.frame(
  Client_Gender = genders,
  Mortality_Rate = sapply(genders, function(gender) {
    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
    calc_mortality_rate_gender(df_gender, "client_gender")
  })
)

# Step 3: Plot Gender vs Mortality Rate as a bar chart
ggplot(mortality_by_gender, aes(x = Client_Gender, y = Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mortality Rate by Gender",
       x = "Client Gender",
       y = "Mortality Rate (%)") +
  theme_minimal() +
  ylim(0, 2)  # Set y-axis limits to 0-100 for percentage

```
# Discussion {#sec-discussion}
